Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-05 TOMS_Paper/assembly.tex
Add COFFEE section

% Quadrature. Tensor. Quadrature <, >, = Tensor
%code evaluating element matrices using quadrature

@@

In this paper, we build on the work presented in~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines. We argue that peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to obtain, for instance, effective register utilization and SIMD vectorization. The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become sub-steps of code specialization; on the other hand, generalized loop-invariant code motion is a step of the expression rewriting process. More importantly, we complement and generalize our previous work with the following contributions.

@@

...TODO...: say that "assembly expression" =THE EXPRE// COMPUTATION OF ELEMENT MATRIX

\subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
In order to generate high performance code, mathematical expressions evaluating the element tensor must be optimized with regards to several interrelated aspects: 1) minimization of floating point operations, 2) instruction-level parallelism, and 3) data locality. In this paper, we tackle these three points building on our previous work (\cite{Luporini}).

...We propose a novel structure of how we think a platform-independent, domain-specific optimizing compiler should look like....

@@

The high-level structure of COFFEE is shown in Figure~\ref{fig:coffee-outline}.

...The Expression Rewriter is a software module implemented in COFFEE dealing with the first point...

...The Code Specializer (henceforth CS) is...
A key point is that the ER has to perform transformations that do not break the code specializer. Having indirections is really dangerous...

\subsection{Summary of Optimizations for Automated Quadrature Representation}
\label{sec:summary-opts}
...TODO... : clearly state, once more, we operate in the regime of automated code generation

To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work on automated code transformations for quadrature representation.

...TODO...

The work in~\cite{Luporini} resulted in the development of COFFEE, a compiler for the optimization of local assembly kernels relying on quadrature representation.

...TODO...

Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.

...TODO...

@@

% Ci sono due goal: minimizzare flops e specializzare

@@

\section{Expression Rewriting}
\label{sec:expr-rewriting}
As summarized in~\ref{sec:summary-opts}, loop-invariant code motion is the key to reduce the computational intensity of a mathematical expression. The Expression Rewriter (henceforth ER) that we have designed and implemented in COFFEE enhances this technique by making two steps forward, which allow more redundant computation to be avoided.

@@

An example is given in Figure~\ref{fig:tovectexpand-code}: \texttt{F1, F2, F3, T1} are all scalar quantities, and their computation will not be auto-vectorized (as explained in Section~\ref{sec:perf-results}, we experimented with the Intel and GNU compilers). The argument that the cost of evaluating such expressions is not relevant because their computational cost is $O(I)$, while the evaluation of the element matrix has a cost of $O(I \cdot T^2)$, is too superficial, in general. There are notable cases, for instance problems based on hyperelasticity, for which the Expression Rewriter produces a significantly-large number of invariant scalar terms, which dramatically impact the execution time.

@@

>>>>

@@

In this paper, we build on the work presented in~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines based on quadrature representation. We argue that peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to obtain, for instance, effective register utilization and SIMD vectorization. The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become sub-steps of code specialization; on the other hand, generalized loop-invariant code motion is a step of the expression rewriting process. More importantly, we complement and generalize our previous work with the following contributions.

@@

In the next sections, we will often refer to the (local) element matrix evaluation, such as Equation~\ref{quadrature} for the weighted Lapalce operator, as the \textit{assembly expression} of the variational problem.

\subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
\label{sec:summary-opts}
If high performance code needs to be generated, assembly expressions must be optimized with regards to three interrelated aspects: 1) arithmetic intensity, 2) instruction-level parallelism, and 3) data locality. In this paper, we tackle these three points building on our previous work (\cite{Luporini}).

@@

COFFEE is a mature, platform-independent compiler capable of optimizing local assembly code. Its high-level structure can be outlined as in Figure~\ref{fig:coffee-outline}. The input is either suitably-annoted C code, explicitly provided by the user, or a decorated abstract syntax tree (henceforth AST) representation of an assembly kernel, automatically generated by the higher level in the stack. We have integrated COFFEE with the Firedrake framework; here, the input to COFFEE is originated by the FEniCS Form Compiler in the form of an AST. Decoration of special AST nodes, or equivalently annotation of C code, is extensively used to convey domain knowledge to COFFEE. This is used to introduce more effective, specialized optimizations, as well as to simplify implementation.

Two conceptually distinct software modules can be individuated: the Expression Rewriter and the Code Specializer. The former targets arithmetic intensity. Its role is to transform the assembly expression and, therefore, the enclosing loop nest, so as to minimize the number of floating point operations that are performed to evaluate the element matrix (or vector). The latter is tailored to optimizing for instruction-level parallelism, particularly SIMD vectorization, and data (register) locality. The Expression Rewriter manipulates and transforms the original AST, which is then provided to the Code Specializer. As described later, we have invested some effort to ensure that the code transformations applied in the expression rewriting stage do not break any code specialization opportunity. This for example would not be the case if the Expression Rewriter coincided with the FEniCS Form Compiler's built-in optimization system; we will clarify this aspect in Section~\ref{sec:zeros}.

To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work. Moreover, in Section~\ref{sec:perf-results}, we will explicitly compare the performance achieved with the transformations presented in this paper to~\cite{Luporini} (as well as to FEniCS' built-in optimizations).

In our previous work, we have demonstrated the effectiveness of a set of optimizations for finite element quadrature-based integration routines originated through automated code generation. We have also shown how different subset of optimizations are needed to achieve close-to-peak performance in different problems. In particular, we can summarize our study as follows

\begin{itemize}
\item \emph{Generalized Loop-invariant Code Motion}. Compiler's loop-invariant code motion algorithms may not be general enough to optimize assembly expressions, in which different sub-expressions are invariant with respect to more than one loop in the enclosing nest. As briefly described in Section~\ref{sec:expr-rewriting}, we work around this limitation, while also achieving vectorization of invariant code.
\item \emph{Padding and data alignment}. The small size of the loop nest require all of the involved arrays to be padded to a multiple of the vector register length so as to maximize the effectiveness of SIMD code. Data alignment can be enforced as a consequence of padding. 
\item \emph{Vector-register Tiling}. Blocking at the level of vector registers, which we perform exploiting the specific memory access pattern of the assembly expressions (i.e. a domain-aware transformation), improves data locality beyond traditional unroll-and-jam optimizations. This is especially true for relatively high polynomial order (i.e. greater than 2) or when pre-multiplying functions are present.
\item \emph{Expression Splitting}. In certain assembly expressions the register pressure is significantly high: when the number of basis functions arrays (or, equivalently, temporaries introduced by loop-invariant code motion) and constants is large, spilling to L1 cache is a consequence for architectures with a relatively low number of logical registers (e.g. 16/32). We exploit sum's associativity to ``split'' the assembly expression into multiple sub-expressions, which are computed individually.
\end{itemize}

In Figure~\ref{fig:coffee-outline}, we show where these transformations are logically applied in COFFEE; also, the contributions of this work are highlighted. In the next two sections, we describe the new functionalities of, respectively, the Expression Rewriter and the Code Specializer.

%Temporary arrays can be placed at the right depth in the sorrounding loop nest to store values of sub-expressions that are invariant to one or more loops.

@@

@@

\section{Expression Rewriting}
\label{sec:expr-rewriting}
As summarized in~\ref{sec:summary-opts}, loop-invariant code motion is the key to reduce the computational intensity of an assembly expression. The Expression Rewriter (henceforth ER) that we have designed and implemented in COFFEE enhances this technique by making two steps forward, which allow more redundant computation to be avoided.

@@

An example is given in Figure~\ref{fig:tovectexpand-code}: \texttt{F1, F2, F3, T1} are all scalar quantities, and their computation will not be auto-vectorized (as explained in Section~\ref{sec:perf-results}, we experimented with the Intel and GNU compilers). The argument that the cost of evaluating such expressions is not relevant because their computational cost is $O(I)$, while the evaluation of the element matrix has a cost of $O(I T^2)$, is too superficial, in general. There are notable cases, for instance problems based on hyperelasticity, for which the Expression Rewriter produces a significantly-large number of invariant scalar terms, which dramatically impact the execution time.

@@

% Quadrature. Tensor. Quadrature <, >, = Tensor
